---
title: "Lecture 3"
author: "Elizabeth Colantuoni"
date: "1/28/2025"
geometry: "left=1in,top=0.75in,right=1in,bottom=0.75in"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages("tidyverse")
# install.packages("ggplot2")
# install.packages("splines")
# install.packages("dplyr")
# install.packages("RColorBrewer")
# install.packages("GGally")
library(RColorBrewer)
library(ggplot2)
library(splines)
library(dplyr)
library(GGally)
library(gridExtra)
```

# I. Introduction

In real estate, there are three principles: “location, location, location”.

In data analysis (empirical science, generally), the corresponding principles are: “question, question, question”.

In Lectures 3 and 4, we will look at three questions about Nepali children’s growth using the Nepal Children’s Anthropomety data kindly provided by Joanne Katz, Professor of International Health and her colleagues.

The questions are:

1. How does the population mean (i.e. average) arm circumference (AC) vary as a function of child’s age? 

2. Is the AC-age relationship the same for boys and girls?

3. Among children of the same height, how does the population mean AC vary as a function of age and is the relationship the same for boys and girls?

We will address Question 1 in Lecture 3 and Questions 2 and 3 in Lecture 4.

# II. The Data

In this section, we will read in the data and perform some basic processing and data visualizations to prepare for the analysis.

## A. Read in and look at the data

```{r readindata}
load("nepal.anthro.rdata")
dim(nepal.anthro)
names(nepal.anthro)
```

\newpage


## B. Analysis sample  

Extract the key variables we need for our analysis and only the first row of data for each child.  NOTE:  The data provides multiple observations over time for each child.  Initially, we will evaluate the first assessment (baseline) for each child.

```{r subsetdata}
d= nepal.anthro %>% select(.,age,sex,ht,wt,arm,num) %>% filter(.,num==1)
d <- d[complete.cases(d),] # drop cases without one or more of these variables
d <- d[order(d$age),-6] # Sort the data by age and drop "num"
dim(d)
head(d)
```

## C. Display key variables

You should ALWAYS visualize your data.  One quick approach is to make pairwise scatterplots where you visualize the association between each pair of variables.

The pairs plot (you find the ggplot version; see ggpairs) is a convenient way to see the pairwise scatterplots in the dataset.

It is a good idea to include the Y and X variables, putting the Y variable last so the bottom row is the plot of Y against each individual X.

```{r pairsplot,fig.height=4,fig.width=6,fig.align='center'}
par(mar=c(1,1,1,1))
pairs(select(d,age,sex,ht,wt,arm),pch=".")
```

For Question 1, we will focus on AC and age.  Here we make a plot to take a closer look the relationship between these two variables.

```{r closerplot,fig.height = 2.5, fig.width = 5, fig.align = "center",warning=FALSE}
ggplot(d, aes(x = age, y = arm)) +
    geom_jitter(alpha = 0.5) +
    theme_bw() + 
    scale_y_continuous(breaks=seq(8,18,2),limits=c(8,18)) +
    scale_x_continuous(breaks=seq(0,60,6),limits=c(0,60)) +
    labs(y = "Arm cirumference (in cm)", x = "Age (in months)") +
    theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"))
```

**Q1: Describe the relationship between AC and age.**

\vspace{1in}



# III. Define, fit and interpret a linear spline model

In this section, we will define the necessary variables for, fit and interpret a linear spline model to the relationship between population mean AC and age.

## A. Define the spline terms for inclusion in model

We will start with knots at 6, 12, 24, 36, 48 months and then drop unnecessary knot points (i.e. where allowing for a change in slope is not necessary).

```{r definesplines}
d=mutate(d,
agesp6=ifelse(age-6>0, age-6,0),
agesp12=ifelse(age-12>0, age-12,0),
agesp24=ifelse(age-24>0, age-24,0),
agesp36=ifelse(age-36>0,age-36,0),
agesp48=ifelse(age-48>0, age-48,0)
)
# check what predictors for linear splines look like
pairs(select(d,age,agesp6,agesp12,agesp24,agesp36, agesp48),pch=".",
main="Pairs Plot of Age Linear Spline Vars")
```

## Fit, interpret and visualize the linear spline regression model

Include all the spline terms at once allowing for 5 slope changes over the 5 years of age, call this model Model 1.

**Q2: Write down the mathematical representation for the model allowing for 5 slope changes over the 5 years of age.**
$y = \beta_0 + \beta_1age_i + \beta_2(age_i-6)^+ + \beta_3(age_i -12)^+ + \beta_4(age_i -24)^+ + \beta_5(age_i -36)^+ + \beta_6(age_i -48)^+ + \epsilon_i$
$\epsilon_i ~ N(0, \sigma^2)$
Example: find mean of Y- when age = 12
$E(Y_i) = \beta_0 + \beta_1(12) + \beta_2(6)^+$

\newpage

```{r model1}
cc=complete.cases(select(d,age,arm))
d.cc=filter(d,cc)
d.cc = arrange(d.cc,age)
reg1<-lm(data=d.cc, arm~age+agesp6+agesp12+agesp24+agesp36+agesp48)
summary.lm(reg1)
```

**Q3: Interpret the value of the intercept**

\vspace{0.5in}

The average arm circumference of a 0 month old child is 10.94 cm. Note, this is an extrapolation because the youngest child in the sample is 1 month old. Confidence interval: $10.94 \pm 1.96*0.55$


**Q4: Interpret the coefficient for "age"**

\vspace{0.5in}
For children under 6 months, the difference in average arm circumference comparing children who differ by one month of age is $\beta_1 + \beta_2$

**Q5: Interpret the coefficient for "agesp6"**

\vspace{0.5in}

For children between 6 and 12 months, for each increase in 1 month of age, the average arm circumference

**Q6: What is the estimate of the population average, standard deviation and variance of AC among 12-month old children?**

\vspace{1in}

\newpage

The figure below overlays the observed data with the estimated population mean AC at each age.

```{r model1fit,fig.height = 2.5, fig.width = 5, fig.align = "center",warning=FALSE}
ggplot(d.cc, aes(x = age, y = arm)) +
  theme_bw() + 
  geom_jitter(alpha = 0.5) +
  geom_line(aes(x = age, y = reg1$fitted.values),lwd=1.25) +
  geom_smooth(span=0.3,lwd=1) +
  scale_y_continuous(breaks=seq(8,18,2),limits=c(8,18)) +
  scale_x_continuous(breaks=seq(0,60,6),limits=c(0,60)) +
  labs(y = "Arm cirumference (in cm)", x = "Age (in months)") +
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"))
```

**Q7: Do you think your Model 1 is a useful model?  Are there any trends in how the estimated population mean AC changes with age that you find unusual or unlikely?**

\vspace{1cm}


\newpage

# IV. Fine-tuning the model

Because there is little reason to think that mean AC goes down between ages 6 and 12 months and because there is little evidence in the data supporting changes in slope after 12 months, let’s fit a second model with only two slope changes at 6 and 12 months, then a third with only a break at 6 months.

```{r models23}
reg2<-lm(data=d.cc, arm~age+agesp6+agesp12)
reg3<-lm(data=d.cc, arm~age+agesp6)
summary.lm(reg2); summary.lm(reg3)
```

**Q8: Without doing a formal statistical test, consider the size and standard error of the estimate for "agesp12" and make a recommendation about the need to allow the AC vs. age relationship to vary comparing children 6 to 12 months of age to children over 12 months of age.**

\vspace{1cm}

\newpage

Make a figure of the observed data and the 3 linear spline models.

```{r model23fit,fig.height = 2.5, fig.width = 5, fig.align = "center",warning=FALSE}
ggplot(d.cc, aes(x = age, y = arm)) + theme_bw() + 
  geom_jitter(alpha = 0.5) +
  geom_line(aes(x = age, y = reg1$fitted.values),color="black") +
  geom_line(aes(x = age, y = reg2$fitted.values),color="red") +
  geom_line(aes(x = age, y = reg3$fitted.values),color="blue") +
  scale_y_continuous(breaks=seq(8,18,2),limits=c(8,18)) +
  scale_x_continuous(breaks=seq(0,60,6),limits=c(0,60)) +
  labs(y = "Arm cirumference (in cm)", x = "Age (in months)") + 
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"))
```

**Q9: How does the population mean AC vary as a function of child’s age?  Write an answer using the results of your favorite model among 1-3. Write in scientific terms, use units, be numerate. This is an exercise to put into your own words the results of a simple regression analysis.**


